Countering language attrition with PanLex and the Web of Data

### 2014-01-22 panlex-rdf-paper/2012/MLOD_PanLex/paper.tex
Added the panlex paper to the repo

%\usepackage{verbatim}
\lstset{%
    numberbychapter=false,
    numbers=left,
    numberstyle=\tiny,
    basicstyle=\footnotesize\ttfamily,
    tabsize=2,
    framexleftmargin=2pt,
    captionpos=b,
    frame=single,
    breaklines=true
}
\lstdefinestyle{rdf}{numberblanklines=true, morekeywords={}}
\lstdefinestyle{sparql}{basicstyle=\ttfamily\scriptsize,numberblanklines=true,
    morekeywords={SELECT,OPTIONAL,FROM,DISTINCT,a,WHERE,FILTER,GROUP,ORDER,LIMIT,BY,IN,AS},
    emph={r,pub,aairObject,verb,person,bday,s,p,o},emphstyle=\textit
}
\lstdefinestyle{turtle}{basicstyle=\ttfamily\tiny ,numberblanklines=true,
    morekeywords={a, @prefix},
    morecomment=[s][\textrm]{<}{>},
    morecomment=[s][\textit]{"}{"},
}

@@

%patrick.westphal@studserv.uni-leipzig.de
\author[1]{\fnms{Patrick} \snm{Westphal}},
\author[2]{\fnms{Claus} \snm{Stadler}},
\author[3]{\fnms{Jonathan} \snm{Pool}}
\address[1]{University of Leipzig, patrick.westphal@studserv.uni-leipzig.de}
\address[2]{University of Leipzig, cstadler@informatik.uni-leipzig.de}
\address[3]{Long Now Foundation, San Francisco, pool@utilika.org}

@@

\begin{abstract}
At present, there are approximately 7,000 living languages in the world.
However, some experts claim that the process of globalization may eventually lead
to the world losing this linguistic diversity.
The vision of the PanLex project is to save these languages, especially low-density ones,
by allowing them to be intertranslatable and thus to be a part of the Information Age.
For this reason, PanLex gathers and integrates information from thousands of
linguistic resources, such as monolingual dictionaries, bilingual dictionaries, multilingual dictionaries,
glossaries, standards and thesauri.
In this paper we describe how we transformed this data to RDF, interlinked it with
Lexvo and DBpedia and published it as Linked Data and SPARQL.
\end{abstract}

@@

\section{Introduction}
% \begin{itemize}
% \item introduce PanLex project
% \item @author: Jonathan
% \item Motivation: About approximately 6900 active languages (\todo{source seems to be http://www.amazon.com/Ethnologue-Languages-M-Paul-Lewis/dp/1556712162}), Existing translation only between relatively few language pairs (dictionaries).
% \item Transitive word-by-word translations 
% * high density and low density languages
% * decline in linguistic diversity
% * These combined forces have led to predictions that something between half and 90%
%   of the world's living languages will die within the next century (Woodbury 2006; UNESCO 2003: 2).
%   Weaker forms of these forces appear to be shrinking the use of medium-density languages in science,
%   diplomacy, business, and other domains (Phillipson 2008).
% * When languages die, they argue, the world loses
%   (1) irreplaceable knowledge of history, medicine, nature, and productive methods encoded in languages'
% * extraction + integration
% \end{itemize}
At present, there are about 7,000 living languages in the world\footnote{See \url{http://www.sil.org/iso639-3/download.asp} for a list of registered languages}.
Nonetheless, some experts claim that processes such as nation-state consolidation and globalization are producing language attrition so rapidly that up to 90\% of all languages alive today will be extinct within a century~\cite{endang_lang}.
Theorists of biolinguistic diversity argue that the loss of
language diversity, the loss of human biological knowledge, and the loss of
species diversity are mutually supportive and thus that language preservation
and revitalization are essential to the preservation of biological diversity~\cite{nettle}.
The vision of the PanLex project is to save all these thousands of languages,
especially those low-density ones that are threatened by extinction,
by supporting their use in global communication. This requires panlingual
translation: translation from any language of the world into any other.
One of the crucial components of panlingual translation of discourses is
panlingual lexical translation. PanLex is designed to support that component.
It documents the known lexical translations (translations of lexemes) among all
languages.

@@

In \autoref{sec:triplificate} we will introduce the PanLex dataset, present our
PanLex RDF vocabulary and explain how we transformed the one into the other.
The following section is about the actual dataset publishing and how we linked
to other datasets of the LLOD cloud.
Finally we discuss related work, conclude our approach and give some hints to future work.

\section{Triplification of the Raw Data}
\label{sec:triplificate}
In this section we first provide an analysis of the original PanLex data.
Subsequently, we introduce our URI and vocabulary design and we explain 
the steps taken to publish the data as RDF.
%The vocabulary 
%afterwards we show how the original data was converted to RDF and published
%via a SPARQL endpoint.
%The interlinking of the created RDF data with other datasets is discussed in the
%following section.

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
At the core, PanLex is a \emph{database}, served by a PostgreSQL database server, that ``represents \emph{assertions} about the \emph{meanings} of \emph{expressions}''\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
As of now, the database contains about 20 million meanings and 19 million expressions extracted from about 2,000 sources.

Assertions are factual claims attributed to a certain source and editor.
They do not represent absolute truths and as such it is valid that different editors make contradictory assertions based on the same sources.

Expressions are lexemes (dictionary entries) and are related to their respective languages.
Furthermore, expressions have meanings. All expressions sharing the same meaning are either translations or synonyms.

The lexical database of the PanLex project is based on different sources like monolingual dictionaries,
bilingual dictionaries, multilingual dictionaries, glossaries, standards, and thesauri.
The list of used sources can also be found on the web\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.

The actual data derived from these sources comprises single- and multi-word expressions and meanings assigned to them.
The most important entities and their relations are depicted in \autoref{fig:db-schema} and are explained in more the detail in the following.

@@

\begin{itemize}
  \item \emph{Expressions} are lemmas. For example, \emph{``go''} is a valid expression, whereas \emph{``went''} is not.
  Expressions are always given in a language variety and can only be given once per language variety.
  They were extracted from the afore-mentioned mono- or bilingual dictionaries or from other sources.
  \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
  \item \emph {Language varieties} within a language are further distinguished by arbitrary unique integer identifiers.
  For example, ``eng'' is the ISO 639-3 code for the English language.
  Panlex defines various varieties of it, including ``English'' (eng-000), ``Simple English'' (eng-001) and ``British English'' (eng-005).
  Their labels, when possible, are autonyms, written in the native writing system.
  So, in contrast to mechanisms like IETF BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}} there is no need for a transcription. %  and one can use variety names like ``{زبان اوستایی}''.
  Additionally these variety identifiers and labels do not have to be registered like the IETF BCP 47 language subtags\footnote{\url{http://www.iana.org/assignments/language-subtag-registry}}.
  \item \emph{Meanings} of expressions are distinguished by editors based on their
  interpretation of an information source.
%  Information sources can be dictionaries, thesauri, etc., and also personal knowledge. 
  The pair comprised of the editor and the information source is referred to as a meaning's \emph{approver}.
  Editors may add informal definitions to meanings in any language variety.
  \item \emph{Definitions} are optional descriptions of a meaning.
  They are given in a certain language variety.
  A description of the verb \emph{browse} for example could be \emph{``move or surf through various files on a computer, the Internet, etc.''}, marked as a definition in the ``English'' language variety.
  \item \emph{Denotations} relate expressions to meanings.
  So homonyms are represented by one expression being connected to different meanings.
  Denotations may be enriched with part-of-speech tags selected from a closed list based on the
  Open Lexicon Interchange Format (OLIF) standard\footnote{\url{http://www.olif.net/}}.
  For example, \emph{fall} can be a verb or a noun for autumn.
  Furthermore, it is possible to generically tag denotations with metadata
  using key-value pairs.
  For instance, an English expression \emph{pig} when synonymous with
  \emph{police officer} could be annotated with \emph{pragmatics=vulgar}.
  \item \emph{Users} have editorial privileges over the language varieties and the approvers that they define.
  \item \emph{Licenses} are also considered by the PanLex project.
  At present there are ten different license categories an approver can be annotated with.
  They are \emph{public domain}, \emph{Creative Commons}, \emph{request} (meaning that one has to ask the author of the resource), \emph{GNU General Public License}, \emph{GNU Lesser General Public License}, \emph{GNU Free Documentation License}, \emph{MIT License}, \emph{copyright} (stating that there is a certain copyright holder), \emph{other} and \emph{unknown}.
\end{itemize}

@@

%The same expression may be mentioned in multiple information sources.
%However, an expression's possible meanings depend on the user's interpretion of them in context of such a source.
%For this reason, every meaning depends on an approver.
%\emph{Denotations} capture the mappings between expressions and their possible meanings.  
Note that \emph{approver} combines a user with an \emph{information source},
however the \emph{information source} is not modelled as a distinct entity.
Also, the information of whether or not two meanings with different approvers
are the same is not being captured.

@@

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
%The schema described above was adopted to create RDF triple structures.
%is a good starting point most of the relations were taken over introducing a PanLex RDF vocabulary.
The entities and relations of the schema described in the previous section
serve as the base for the development of the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside
in the namespace \texttt{\small <http://panlex.org/plx/>}, abbreviated with \texttt{\small plx}.
An example of the resulting ontology is depicted in \autoref{fig:vocabulary} and
summarized as follows:
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern
\emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case
and the \{id\} is the primary key of the corresponding database table.

%As in the original PanLex database there are also \emph{expressions} used in the RDF dataset.
\begin{itemize}
\item Expressions are modeled as instances of \\ \texttt{\small plx:Expression}. % manual linebreak
Their textual representations and normalized versions thereof become the 
values of the properties \texttt{\small rdfs:label} and \\ \texttt{\small plx:degradedText}, respectively. % manual linebreak
Their corresponding language variety is stated using \\ \texttt{\small plx:languageVariety}. % manual linebreak

\item For language and language varieties the classes
\texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
\emph{ISO 639-1} and \emph{ISO 639-3 codes} are of the classes
\texttt{\small plx:Iso639-1Code} and \\ \texttt{\small plx:Iso639-3Code} respectively. % manual linebreak

\item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
Entities of this class may have an identifier assigned with the
\texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
%This meaning identifier represents ids given to the considered meaning in other
%data sources and may at a later stage be used to build links to the meaning
%as represented on these sites.
Meanings may also have \emph{definitions}, entities of the
\texttt{\small plx:Definition} class, giving a textual representation
(\texttt{\small rdfs:label}) with more than just a lexeme in a certain
language variety (\texttt{\small plx:languageVariety}).

%These resources have an \texttt{rdfs:label} containing the textual representation of the considered expression.
%Additionally there is a normalized version of the label, reachable via the \texttt{plx:degradedText} property.
%Expression entities also have an assigned language variety (\texttt{plx:languageVariety}).
%It refers to the language variety the textual representation is written in.
\item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
These are entities of the \texttt{\small plx:Denotation} class pointing to meanings and expressions via the \texttt{\small plx:denotationMeaning} and \\ \texttt{\small plx:denotationExpression} property respectively. % manual linebreak
Denotations may also have a word class assigned to them.
This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
\item All approvers share the \texttt{\small plx:Approver} class. The characteristics of an approver are described using mainly triples with literal objects.
These are for example \texttt{\small plx:registrationDate} pointing to an \texttt{\small xsd:date} literal,
\texttt{\small dc:title} to assign the title of a source,
\texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name.
\end{itemize}

@@

Since the PanLex project compiled its database by extracting data from different sources, the licenses of these sources were also considered.
At present, we support the different license categories given in the database by creating resources of the \texttt{\small plx:License} class.
%with appropriate descriptions in their \texttt{rdfs:label}.
%In the RDF dataset these have an additional \texttt{rdfs:label} property giving the names or descriptions 
\begin{table}
\begin{scriptsize}
\begin{tabular}{ll}
Class & Properties \\
\toprule
\texttt{plx:Approver} & \texttt{plx:registrationDate} \\
                      & \texttt{rdfs:label} \\
                      & \texttt{foaf:homepage} \\
                      & \texttt{plx:isbn} \\
                      & \texttt{dc:creator} \\
                      & \texttt{dc:title} \\
                      & \texttt{dc:publisher} \\
                      & \texttt{plx:yearOfPublication} \\
                      & \texttt{plx:quality} \\
                      & \texttt{plx:license} \\
\midrule
\texttt{plx:Language}
                      & \texttt{plx:iso639-3Code} \\
                      & \texttt{plx:iso639-1Code} \\
\midrule
\texttt{plx:LanguageVariety}
                      & \texttt{plx:languageVarietyOf} \\
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:Iso639-1Code} & \\
\midrule
\texttt{plx:Iso639-3Code} & \\
\midrule
\texttt{plx:Expression}
                      & \texttt{plx:languageVariety} \\
                      & \texttt{rdfs:label} \\
                      & \texttt{plx:degradedText} \\
\midrule
\texttt{plx:Meaning}
                      & \texttt{plx:approver} \\
                      & \texttt{plx:meaningDefinition} \\
                      & \texttt{plx:identifier} \\
\midrule
\texttt{plx:Definition}
                      & \texttt{plx:languageVariety} \\
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:Denotation}
                      & \texttt{plx:denotationMeaning} \\
                      & \texttt{plx:denotationExpression} \\
                      & \texttt{plx:wordClass} \\
\midrule
\texttt{plx:WordClass}  
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:License}  & \texttt{rdfs:label} \\
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Classes and properties used in the PanLex RDF vocabulary. Note that all \texttt{rdf:type} and \texttt{rdfs:label} properties have been omitted for brevity.}
\label{tbl:vocabulary}
\end{table}

%An overview of the PanLex vocabulary is shown in \autoref{fig:vocabulary}.%
\begin{figure}
\includegraphics[width=0.47\textwidth]{images/pdf/ontology.pdf}
\caption{Overview of the PanLex RDF vocabulary}
\label{fig:vocabulary}
\end{figure}

>>>>

%\usepackage{verbatim}
\lstset{%
    numberbychapter=false,
    numbers=left,
    numberstyle=\tiny,
%    basicstyle=\footnotesize\ttfamily,
    basicstyle=\ttfamily \tiny,
    tabsize=2,
    framexleftmargin=2pt,
    captionpos=b,
    frame=single,
    breaklines=true
}
\lstdefinestyle{rdf}{numberblanklines=true, morekeywords={}}
\lstdefinestyle{sparql}{basicstyle=\ttfamily\scriptsize,numberblanklines=true,
    morekeywords={SELECT,OPTIONAL,FROM,DISTINCT,a,WHERE,FILTER,GROUP,ORDER,LIMIT,BY,IN,AS},
    emph={r,pub,aairObject,verb,person,bday,s,p,o},emphstyle=\textit
}
\lstdefinestyle{turtle}{basicstyle=\ttfamily\tiny ,numberblanklines=true,
    morekeywords={a, @prefix},
    morecomment=[s][\textrm]{<}{>},
    morecomment=[s][\textit]{"}{"},
}

@@

%patrick.westphal@studserv.uni-leipzig.de
\author[A]{\fnms{Patrick} \snm{Westphal}},
\author[A]{\fnms{Claus} \snm{Stadler}},
\author[B]{\fnms{Jonathan} \snm{Pool}}
%\address[1]{University of Leipzig, patrick.westphal@studserv.uni-leipzig.de}
\address[A]{University of Leipzig, \{pwestphal, cstadler\}@informatik.uni-leipzig.de}
\address[B]{Long Now Foundation, San Francisco, pool@panlex.org}

@@

\begin{abstract}
At present, there are approximately 7,000 living languages in the world.
However, some experts claim that the process of globalization may eventually lead
to the world losing this linguistic diversity.
The vision of the PanLex project is to help save these languages, especially
low-density ones, by allowing them to be intertranslatable and thus to be a part of the Information Age.
For this reason, PanLex gathers and integrates information from thousands of
linguistic resources, such as monolingual dictionaries, bilingual dictionaries, multilingual dictionaries,
glossaries, standards and thesauri.
In this dataset description paper we detail how we transformed this data to
RDF, interlinked it with Lexvo and DBpedia and published it as Linked Data and
via SPARQL.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
% \begin{itemize}
% \item introduce PanLex project
% \item @author: Jonathan
% \item Motivation: About approximately 6900 active languages (\todo{source seems to be http://www.amazon.com/Ethnologue-Languages-M-Paul-Lewis/dp/1556712162}), Existing translation only between relatively few language pairs (dictionaries).
% \item Transitive word-by-word translations 
% * high density and low density languages
% * decline in linguistic diversity
% * These combined forces have led to predictions that something between half and 90%
%   of the world's living languages will die within the next century (Woodbury 2006; UNESCO 2003: 2).
%   Weaker forms of these forces appear to be shrinking the use of medium-density languages in science,
%   diplomacy, business, and other domains (Phillipson 2008).
% * When languages die, they argue, the world loses
%   (1) irreplaceable knowledge of history, medicine, nature, and productive methods encoded in languages'
% * extraction + integration
% \end{itemize}
At present, there are about 7,000 living languages in the world\footnote{See \url{http://www.sil.org/iso639-3/download.asp} for a list of registered languages}.
Nonetheless, some experts claim that processes such as nation-state consolidation and globalization are producing language attrition so rapidly that up to 90\% of all languages alive today will be extinct within a century~\cite{endang_lang}.
Theorists of biolinguistic diversity argue that the loss of
language diversity, the loss of human biological knowledge, and the loss of
species diversity are mutually supportive and thus that language preservation
and revitalization are essential to the preservation of biological diversity~\cite{nettle}.
The vision of the PanLex project is to help save these thousands of
languages, especially those low-density ones that are threatened by extinction,
by supporting their use in global communication. This requires panlingual
translation: translation from any language of the world into any other.
One of the crucial components of panlingual translation of discourses is
panlingual lexical translation. PanLex is designed to support that component.
It documents the known lexical translations (translations of lexemes) among all
languages.

@@

In \autoref{sec:triplificate} we introduce the PanLex dataset, present our
PanLex RDF vocabulary and explain how we transformed the one into the other.
\autoref{sec:linking} is about how we linked to other datasets of the LLOD
cloud, whereas \autoref{sec:publishing} is about the actual dataset publishing.
Usage scenarios are given in \autoref{sec:usage}.
In \autoref{sec:related} we discuss related work, and finally, in
\autoref{sec:conclusion} we conclude our approach and give some hints to future
work.

\section{Triplification of the Raw Data}
\label{sec:triplificate}
In this section we first provide an analysis of the original PanLex dataset.
Subsequently, we introduce our URI and vocabulary design and we explain 
the steps taken to publish the data as RDF.
%The vocabulary 
%afterwards we show how the original data was converted to RDF and published
%via a SPARQL endpoint.
%The interlinking of the created RDF data with other datasets is discussed in the
%following section.

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
At the core of the PanLex \emph{project}, there is the PanLex
\emph{database} which is created from the imports of thousands of lexical
resources, such as monolingual dictionaries, bilingual dictionaries, multilingual dictionaries, glossaries, standards, and thesauri.
The concrete list of used sources is available online\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.
The data derived from these sources comprises single- and multi-word expressions and meanings assigned to them.
Conceptually, the PanLex database thus ``represents \emph{assertions} about the \emph{meanings} of \emph{expressions}''\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
As of now, the database contains about 20 million meanings and 19 million expressions extracted from about 2,000 sources.
The most important entities and their relations are depicted in \autoref{fig:db-schema} and are explained in more the detail in the following.

%At the core, PanLex is a \emph{database}, served by a PostgreSQL database
%server, that ``represents \emph{assertions} about the \emph{meanings} of
% \emph{expressions}''\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
%As of now, the database contains about 20 million meanings and 19 million
% expressions extracted from about 2,000 sources.

%Assertions are factual claims attributed to a certain source and editor.
%They do not represent absolute truths and as such it is valid that different
% editors make contradictory assertions based on the same sources.

%Expressions are lexemes (dictionary entries) and are related to their
% respective languages.
%Furthermore, expressions have meanings. All expressions sharing the same
% meaning are either translations or synonyms.

%The lexical database of the PanLex project is based on different sources like
% monolingual dictionaries, bilingual dictionaries, multilingual dictionaries, glossaries, standards, and thesauri.
%The list of used sources can also be found on the
% Web\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.

%The actual data derived from these sources comprises single- and multi-word
% expressions and meanings assigned to them.
%The most important entities and their relations are depicted in
% \autoref{fig:db-schema} and are explained in more the detail in the following.

@@

\begin{itemize}
  %\item A meaning's \emph{approver} is a pair comprised of the editor and the considered information source.
  \item The starting point of the data acquisition is the \emph{approver}
  entity: An editor processes the content of a certain
  mono- or multilingual source as mentioned above and adds it to the PanLex dataset.
  This combination of a user and a source is referred to as an approver.

  \item \emph{Expressions} are lemmas, i.e. dictionary entries. For example,
  \emph{``go''} is a valid expression, whereas \emph{``went''} is not.
  Expressions are always given in a language variety and can only be given once per language variety.
  %They were extracted from the afore-mentioned mono- or bilingual dictionaries
  % or from other sources.
  \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
  \item \emph {Language varieties} allow one to make more fine grained
  distinctions within a language. Their codes are composed of the language code
  combined with a PanLex specific identifier.
  For example, ``eng''
  is the ISO 639-3 code for the English language.
  Panlex defines various varieties of it, including ``English'' (eng-000), ``Simple English'' (eng-001) and ``British English'' (eng-005).
  Their labels, when possible, are autonyms, written in the native writing system.
  So, in contrast to mechanisms like IETF BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}} there is no need for a transcription. %  and one can use variety names like ``{زبان اوستایی}''.
  %Additionally these variety identifiers and labels do not have to be registered like the IETF BCP 47 language subtags\footnote{\url{http://www.iana.org/assignments/language-subtag-registry}}.
  \item \emph{Meanings} in PanLex are entities of which each represents a unique
  possible sense of an expression. 
  %They are allocated as necessary by an editor based on his or her
 %interpretation of expressions.
  Meanings are assigned by editors based on their interpretation of expressions.
  Usually this assignment is done on a per-source basis so that identical meanings
  across multiple sources are not resolved.
  %different
  % meaning identifiers may refer to the same intended sense.
  This means that if there is e.g. a translation of the fruit \emph{``apple''}
  in an English to German dictionary and another translation from English to
  French, these do not necessarily result in a single meaning entity linking to
  all involved languages.
  %the to corresponding expressions in English, German and French.
  Instead, there could be \emph{``apple''} and \emph{``Apfel''} sharing one
  meaning entity and \emph{``apple''} and \emph{``pomme''} sharing another.
%  Editors may add informal definitions to meanings in any language variety.
%  Information sources can be dictionaries, thesauri, etc., and also personal knowledge. 
%  The pair comprised of the editor and the information source is referred to as a meaning's \emph{approver}.
  \item \emph{Definitions} are optional descriptions of a meaning.
  They are given in a certain language variety.
  A description of the verb \emph{browse} for example could be \emph{``move or surf through various files on a computer, the Internet, etc.''}, marked as a definition in the ``English'' language variety.
  \item \emph{Denotations} are entities that relate expressions to meanings and
  may optionally carry annotations in form of sets of key value pairs. For
  instance, an English expression \emph{pig}, when referring to
  \emph{police officer}, could be annotated with \emph{pragmatics=vulgar}. 
  Furthermore, denotations can be tagged with part-of-speech tags, such as word
  classes, selected from a closed list based on the Open Lexicon Interchange Format (OLIF) standard\footnote{\url{http://www.olif.net/}}.
  For example, \emph{fall} can be a verb or a noun for autumn. 
  Homonyms are those expressions that are connected to multiple meanings.
  \item \emph{Users} have editorial privileges over the language varieties and the approvers that they define.
  \item \emph{Licenses} are also considered by the PanLex project.
  At present there are ten different license categories an approver can be annotated with.
  They are \emph{public domain}, \emph{Creative Commons}, \emph{request} (meaning that one has to ask the author of the resource), \emph{GNU General Public License}, \emph{GNU Lesser General Public License}, \emph{GNU Free Documentation License}, \emph{MIT License}, \emph{copyright} (stating that there is a certain copyright holder), \emph{other} and \emph{unknown}.
\end{itemize}

@@

%The same expression may be mentioned in multiple information sources.
%However, an expression's possible meanings depend on the user's interpretion of them in context of such a source.
%For this reason, every meaning depends on an approver.
%\emph{Denotations} capture the mappings between expressions and their possible meanings.  
Note that \emph{approver} combines a user with an \emph{information source},
however the \emph{information source} is not modeled as a distinct entity.
Also, the information of whether or not two meanings with different approvers
are the same is not being captured.

@@

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
%The schema described above was adopted to create RDF triple structures.
%is a good starting point most of the relations were taken over introducing a PanLex RDF vocabulary.
The entities and relations of the schema described in the previous section
serve as the base for the development of the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside
in the namespace \texttt{\small <http://ld.panlex.org/plx/>}, abbreviated with
\texttt{\small plx}.
An example of the resulting ontology is depicted in \autoref{fig:vocabulary} and
summarized as follows:
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern
\emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case
and the \{id\} is the primary key of the corresponding database table.

%As in the original PanLex database there are also \emph{expressions} used in the RDF dataset.
\begin{itemize}
\item Expressions are modeled as instances of the class \texttt{\small plx:Expression}.
Their original and normalized textual representations become the 
values of the properties \texttt{\small rdfs:label} and \texttt{\small plx:degradedText}, respectively.
Their corresponding language variety is stated
using \texttt{\small plx:languageVariety}.

\item For language and language varieties the classes
\texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
\emph{ISO 639-1} and \emph{ISO 639-3 codes} become instances of the classes
\texttt{\small plx:Iso639-1Code} and \texttt{\small plx:Iso639-3Code}.

\item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
Entities of this class may have an identifier assigned with the
\texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
%This meaning identifier represents ids given to the considered meaning in other
%data sources and may at a later stage be used to build links to the meaning
%as represented on these sites.
Meanings may also have \emph{definitions}, entities of the
\texttt{\small plx:Definition} class, giving a textual representation
(\texttt{\small rdfs:label})
%with more than just a lexeme
in a certain
language variety (\texttt{\small plx:languageVariety}).

%These resources have an \texttt{rdfs:label} containing the textual representation of the considered expression.
%Additionally there is a normalized version of the label, reachable via the \texttt{plx:degradedText} property.
%Expression entities also have an assigned language variety (\texttt{plx:languageVariety}).
%It refers to the language variety the textual representation is written in.
\item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
These are entities of the \texttt{\small plx:Denotation} class pointing
to meanings and expressions via the properties \texttt{\small plx:denotationMeaning}
and \\ \texttt{\small plx:denotationExpression}. % manual linebreak
Denotations may also have a word class assigned to them.
This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
\item All approvers share the \texttt{\small plx:Approver} class. The characteristics of an approver are described using mainly triples with literal objects.
These are for example %\texttt{\small plx:registrationDate} pointing to an \texttt{\small xsd:date} literal,
\texttt{\small dc:title} to assign the title of a source,
\texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name.
\end{itemize}

@@

Since the PanLex project compiled its database by extracting data from different sources, the licenses of these sources were also considered.
At present, we support the different license categories given in the database by creating resources of the \texttt{\small plx:License} class.
%with appropriate descriptions in their \texttt{rdfs:label}.
%In the RDF dataset these have an additional \texttt{rdfs:label} property giving the names or descriptions 
\begin{table}
\begin{tiny}
\begin{tabular}{ll}
Class & Properties \\
\toprule
\texttt{plx:Approver} & \texttt{plx:registrationDate}, \texttt{rdfs:label}\\
                      & \texttt{foaf:homepage}, \texttt{plx:license}, \texttt{dc:date}\\
                      & \texttt{dc:creator}, \texttt{dc:title}, \texttt{plx:quality} \\
                      & \texttt{dc:publisher}, \texttt{dbpedia-owl:isbn} \\
\midrule
\texttt{plx:Language}
                      & \texttt{plx:iso639-3Code} \\
                      & \texttt{plx:iso639-1Code} \\
\midrule
\texttt{plx:LanguageVariety}
                      & \texttt{plx:languageVarietyOf}, \texttt{rdfs:label} \\
\midrule
\texttt{plx:Iso639-1Code} & \\
\midrule
\texttt{plx:Iso639-3Code} & \\
\midrule
\texttt{plx:Expression}
                      & \texttt{plx:languageVariety} \\
                      & \texttt{plx:degradedText}, \texttt{rdfs:label} \\
\midrule
\texttt{plx:Meaning}
                      & \texttt{plx:approver}, \texttt{plx:identifier} \\
                      & \texttt{plx:meaningDefinition} \\
\midrule
\texttt{plx:Definition}
                      & \texttt{plx:languageVariety}, \texttt{rdfs:label} \\
\midrule
\texttt{plx:Denotation}
                      & \texttt{plx:denotationMeaning} \\
                      & \texttt{plx:denotationExpression} \\
                      & \texttt{plx:wordClass} \\
\midrule
\texttt{plx:WordClass}  
                      & \texttt{rdfs:label} \\
\midrule
\texttt{plx:License}  & \texttt{rdfs:label} \\
\bottomrule
\end{tabular}
\end{tiny}
\caption{Classes and properties used in the PanLex RDF vocabulary. Note that all \texttt{rdf:type} properties are omitted for brevity.}
\label{tbl:vocabulary}
\end{table}

%An overview of the PanLex vocabulary is shown in \autoref{fig:vocabulary}.%
\begin{figure}
\includegraphics[width=0.46\textwidth]{images/pdf/ontology.pdf}
\caption{Overview of the PanLex RDF vocabulary}
\label{fig:vocabulary}
\end{figure}

\subsection{Transformation workflow}
\label{sec:conversion}
New sources are added to PanLex on almost a daily basis.   
The size of the PanLex database (including indexes) is currently approximately
18 GB, which is big enough to make recurrent RDF conversion cumbersome:
The RDF file of a full conversion of a database is much larger than the
database itself.
Using conventional hardware, it takes impractically long
to convert all the data. This makes
testing and debugging and fixing issues in the modeling or data conversion of the
data very time-consuming.
As the PanLex data already resides in a relational database, the use of a virtual
RDB2RDF\footnote{\url{http://www.w3.org/2001/sw/wiki/RDB2RDF}}  mapping solution is a natural choice.
The \emph{Sparqlify system}\footnote{\url{https://github.com/AKSW/Sparqlify}} offers,
besides an efficient query rewriting engine, also a very easy-to-use mapping
language, called \emph{Sparqlification Mapping Language} (SML).
Essentially, these mappings consist of three clauses:
\begin{itemize}
  \item \emph{From:} Specifies the logical SQL table (i.e. table, view or query) underlying the SML view.
  \item \emph{With:} Defines SPARQL variables by means of expressions over relational columns that yield RDF terms.
  \item \emph{Construct:} The template (i.e. the set of triple patterns) to be generated in this view, based on the SPARQL variable definitions.
\end{itemize}
\autoref{fig:ex:sparqlify-ml} shows an example of an SML view for the
languages in PanLex: From each row of the \emph{i1} table three resources are
created from the \emph{iso3} column and 
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language} in the PanLex and the \emph{schema.org} namespace.
This view-based approach also demonstrates that changing the vocabulary or
adding support for new ones does not require an extract transform load (ETL) process, and can therefore be done with little effort.
%\todo{Maybe we should link to lexvo directly.}
%PanLex's languages  iso 639-3 code 
%typed with plx:Language
%resources are created for
%the languages in PanLex, linked with \emph{lexvo}, and
